A Detection of the Environmental Dependence of the Sizes and Stellar Haloes of Massive Central Galaxies

### 2017-07-31 draft/paper2_v4.tex
New version

    In \citet{hscMassiveI}, we use deep ($>28.5$ \sb{} in $i$-band), high-quality 
    (0.6\asec seeing) images of ${\sim}3000$ massive (\mstar{}$\geq 11.5$) central 
    galaxies at $0.3 < z < 0.5$ from the Hyper Suprime-Cam Subaru Strategic Program 
    to show that more massive central galaxies display more prominent stellar haloes 
    with shallower outer stellar mass density profile (\mden) and higher ellipticity. 
    In this work, we investigate the environment dependence of stellar mass 
    distributions of massive central galaxies using the same dataset.
    Treating stellar mass within a 100 kpc elliptical aperture (\mtot{}) as proxy of 
    total stellar mass, we show that at fixed \mtot{}, massive central galaxies in 
    \mhalo{}$\geq 10^{14}$\msun{} haloes show slightly shallower \mden{} profiles, 
    larger half-light radius, and lower fraction of stellar mass within inner 10 kpc 
    (\minn{}) than the centrals of \mhalo{}$< 10^{14}$\msun{} haloes. 
    At fixed \minn{}--a tracer of the \textit{in situ} stellar mass, central galaxies 
    from more massive haloes host much more significant outer stellar haloes despite 
    their inner \mden{} profiles are very similar with the ones in less massive haloes.
    These results clearly demonstrate that mass--size relation for central galaxies 
    indeed depends on environment at high-\mtot{} end. 
    We also show that the \mtot{}--\minn{} relation can reflect this dependence 
    better and can help diagnose the assembly of massive galaxies and their haloes.
    These results suggest more massive haloes favor more efficient growth of stellar 
    haloes in their central galaxies, which is broadly consistent with the two phase 
    formation scenario of massive galaxies in which their outer envelopes are 
    accumulated through accretions of small systems.

@@

    A key discovery in the last 
    decade has been the dramatic structural transformation of massive quiescent 
    galaxies \citep[e.g.][]{Trujillo2006, vanDokkum2008, Cimatti2008, Damjanov2009, 
    vanderWel2011, Szomoru2012, Patel2013} from $z \approx 2$ to the present day. 
    These observations suggest that the progenitors of $z{\sim} 0$ massive early-type 
    galaxies (ETGs) need to increase their effective radii ($R_{\rm e}$) by a factor 
    of 2--4 over a time span of 10 Gyrs (e.g. \citealt{Newman2012, vdWel2014}). 
    This observational result spurred the development of the ``two-phase'' formation
    scenario for massive ETGs (e.g. \citealt{Oser2010, Oser2012}),
    in which galaxies form a compact central region at $z\sim 2$ through highly 
    dissipative processes (e.g. gas-rich merger or cold gas-accretion;
    \citealt{Hopkins2008, Dekel2009}). 
    They subsequently assemble extended stellar haloes via dry mergers 
    \citep[e.g,][]{Naab2006, Khochfar2006, Oser2010, Oser2012} which can cause 
    significant size growth at late times. 
    An alternative explanation for size growth, progenitor bias, hypothesizes that 
    larger ETGs were quenched more recently, but this explanation is still under 
    active debate (e.g. \citealt{Newman2012, Carollo2013, Poggianti2013, Belli2015,
    Keating2015, Fagioli2016}).

    There have been multiple observational attempts to test the two-phase 
    formation scenario using galaxies at low redshift, by investigating surface 
    brightness or mass density profiles (e.g. \citealt{Huang2013a, Huang2013b, 
    Oh2017}, optical color gradients (e.g. \citealt{LaBarbera2010, LaBarbera2012}), 
    and stellar population gradients \citep[e.g.,][]{Coccato2010, Coccato2011, 
    Greene2015, Barbosa2016}. 
    These observations are generally consistent with the two-phase scenario, 
    but this picture still face many challenges.
    One of the most outstanding issues is the question of whether or not the 
    fraction of accreted stellar mass displays a dependency on dark matter halos mass.

    According to the two-phase formation scenario, the late time growth of massive 
    ETGs is intrinsically tied to the growth of their host dark matter haloes 
    (e.g. \citealt{Leauthaud2012, Behroozi2013, Shankar2013}). 
    Massive ETGs living at the centers of more massive haloes should be subject to 
    higher merger rate compared to ETGs with similar \mstar{} but living in less 
    massive dark matter haloes. 
    A natural prediction of this picture is therefore that the structures of 
    massive ETGs should depend on their ``environment''\footnote{There are 
    multiple definitions of ``environment'' in the literature.  
    In this work, we use ``environment'' and halo mass interchangeably.}.
    Theoretical models based on the two-phase formation scenario predict an 
    environmental dependence of the stellar mass--effective radius relation for 
    ETGs (\mstar{}--$R_{\rm e}$; 
    e.g., \citealt{Shen2003, Guo2009}): the sizes of massive ETGs in more massive 
    haloes should be larger than those in less massive haloes at fixed \mstar{} 
    (e.g. \citealt{Shankar2013, Shankar2014}). 
    Yet, solid evidence of this effect at low redshift is still missing 
    (e.g. \citealt{Nair2010, HCompany13}; but also see \citealt{Yoon2017}).

@@

    Testing this prediction requires deep imaging of ETGs in order to probe the 
    accreted component (the stellar halo) which can extend to 100 kpc and 
    beyond \citep[eg.,][]{hscMassiveI}. 
    This is very challenging observation as the surface brightness profiles of 
    massive ETGs decline rapidly with typical values of $\mu > 26.0$ \sb{} in 
    $i$-band at 100 kpc and at $z\sim0.3$. 
    In \citet[][Paper I hereafter]{hscMassiveI}, we showed that deep, 
    multi-band imaging from the Subaru Strategic Program (SSP; \citealt{HSC-SSP,
    HSC-DR1}) using Hyper Suprime-Cam (HSC; \citealt{Miyazaki2012}, 
    Miyazaki in~prep.) allows us to extract robust surface stellar
    mass density (\mden{}) profiles for {\it individual} galaxies with 
    \logms{}$>11.4$ at $0.3 < z < 0.5$ and out to 100 kpc. 
    In Paper I, we characterized the stellar mass distributions and mass density 
    profiles of stellar haloes of massive ETGs, and showed that there is a large 
    intrinsic scatter in the \mden{} stellar halos of massive galaxies on 100 kpc 
    scales, and that the fraction of accreted stars residing in the stellar halos 
    increases as a function of \mstar{}. 
    In this follow-up paper, we now address a fundamental prediction of the 
    two-phase formation scenario and conclusively show that the extended stellar 
    halos of massive central galaxies depend on dark matter halo mass. 
    In other terms, we reveal the halo mass dependence of the mass-size relation 
    for massive ETGs.

@@

    Magnitudes use the AB system (\citealt{Oke1983}), and are corrected for Galactic 
    extinction using calibrations from \citet{Schlafly11}.
    In this work, we assume $H_0$ = 70~km~s$^{-1}$ Mpc$^{-1}$, ${\Omega}_{\rm m}=0.3$, 
    and ${\Omega}_{\rm \Lambda}=0.7$.
    Stellar mass is denoted \mstar{} and has been derived using a Chabrier Initial Mass 
    Function (IMF; \citealt{Chabrier2003}). Halo mass is defined as $M_{\rm 200b}$ as 
    $M_{\rm 200b}\equiv M(<r_{\rm 200b})=200\bar{\rho} 
    \frac{4}{3}\pi r_{\rm 200b}^3$ where $r_{\rm 200b}$
    is the radius at which the mean interior density is equal to 200 times
    the mean matter density ($\bar{\rho}$). 
    As in \citet{hscMassiveI}, we do not attempt to decompose or distinguish any  
    potential ``intra-cluster'' 
    component (ICL; e.g. \citealt{Carlberg1997, Lin2004, Gonzalez2005, Mihos2005}).

@@

%% ------------------------------------------------------------------------------------ %% 
\subsection{The Role of Environment in the ``Two--Phase'' Formation of 
            Massive Central Galaxies}

    Results presented here are broadly consistent with the ``two-phase'' formation
    scenario for massive central galaxies in a hierarchical universe. 
    Under this picture, intense dissipative processes at $z > 2$ are responsible for 
    the formation of the \textit{in situ} stars in massive central galaxies. 
    After the star formation was quenched, a more gradual second phase is dominated 
    by non-dissipative mergers of satellite galaxies. 
    Since dry minor mergers can effectively deposit \textit{ex situ} stars in the 
    outskirts of massive galaxies (e.g. \citealt{Oogi2013, Bedorf2013}), 
    it is expected that the stellar mass distributions of massive galaxies 
    show halo mass dependence, as the central galaxies of more massive haloes should
    experience more mergers.

    \citet{Shankar2014} investigates the environment dependence of galaxy size using 
    a series of hierarchical semi-analytic models. 
    All models suggest that, at fixed stellar mass, the median size of galaxies 
    strongly increases with halo mass. 
    Although the centrals of more massive halos only experience slightly higher 
    number of major mergers (mass ratio $>1$:3; e.g. \citealt{Hirschmann2013}), 
    the number of minor mergers strongly depend on the assumed dynamical friction time. 
    When short dynamical friction time is favored (e.g. \citealt{Newman2012}), 
    centrals of more massive haloes can have up to four times more minor mergers 
    (1:100--1:3). 
    These extra minor mergers alone can lead to moderate halo mass dependence of 
    galaxy size. 
    Similarly, \citet{Yoon2017} confirms these theoretical predictions using 
    semi-analytic model from \citet{Guo2011}.

    These predictions are in general agreement with our findings here as we do see 
    differences of \mden{} profiles in the outskirts are a major driver of the 
    halo mass dependence, especially for the \minn{}--matched samples. 
    At the same same, whether or not the difference in number of mergers can explain 
    all the environment dependence is yet to be examined.

    Both semi-analytic models and hydrodynamic simulations still have difficulties 
    reproducing the observed mass--size relation for massive galaxies. 
    Besides the reasons related to physical recipes, it is not straightforward for 
    simulations to predict galaxy size in the same way with observations may also 
    contribute to the problem. 
    Since the definition of apertures mass is more clear, it would be interesting 
    to compare with simulations using the \mtot{}--\minn{} relation instead of the 
    mass--size one.

@@

%% ------------------------------------------------------------------------------------ %% 
\subsection{Comparisons with Mass--Size Relations from Other Works}

@@

    Previous works on this topic at lower redshift mostly find no environment dependence
    of mass--size relation (e.g. \citealt{Nair2010, Maltby2010, Cappellari2013, 
    HCompany13}).  
    The discrepancy between these results and ours should be explained by limitation of 
    shallower images, method for estimating galaxy size, differences in \mstar{} and 
    \mhalo{} ranges. 
    \citet{Cerbrian2014} finds that ETGs with \logms{}$>11.5$ are slightly larger in 
    more massive haloes, but they also shows that this environment trend is reversed 
    at lower \mstar{}. 
    \citet{Yoon2017} carefully studies a large sample of $z\sim0.1$ SDSS ETGs using  
    non-parametric method and recovers a clear environment dependence of mass-size 
    relation at \logms{}$>11.2$. 
    They conclude that, at the same \mstar{}, massive ETGs from dense environment 
    are 20--40\% larger than the ones in underdense environment. 
    The definition of environment, estimates of \mstar{} and sizes in \citet{Yoon2017} 
    are different with this work, but the results generally agree with ours.

    Although recent results start to agree that mass--size relation of massive
    ETGs show environment dependence, the situation at higher redshift is complicated. 
    While certain works find environment dependence of mass--size relation for massive 
    quiescent or early-type galaxies that is qualitatively consistent with the 
    low-$z$ results (e.g. \citealt{Papovich2012, Lani2013, Strazzullo2013, 
    Delaye2014}), there are still observations that suggest otherwise (e.g. 
    \citealt{Rettura2010, Raichoor2012, Kelkar2015, Allen2015}). 
    Some of this confusion can be explained by smaller sample size, special environment
    (high-$z$ massive cluster), or impact of massive star-forming galaxies. 
    But it also reflects the intrinsic difficulty of measuring \mstar{} and size 
    of high-$z$ galaxies (e.g. \cite{Price2017}).
    As we suggested earlier, with the help of deeper observations at higher redshift, 
    it would be beneficial to directly compare the rest-frame surface brightness 
    or the stellar mass density profiles instead of using mass--size relation alone 
    (e.g. \citealt{Szomoru2012, Patel2013, Buitrago2017, Hill2017}).
%% ------------------------------------------------------------------------------------ %%

>>>>

    We use ${\sim}100$ deg$^2$ of deep ($>28.5$\sb{} in $i$-band), high-quality 
    (median 0.6\asec seeing) imaging data from the Hyper Suprime-Cam (HSC) survey 
    to investigate the halo mass dependence of the surface mass density profiles 
    and outer stellar envelopes of massive galaxies. 
    The depth of the HSC survey reaches ${\sim}4$ magnitudes fainter than SDSS and 
    enables us to directly trace stellar mass distributions to 100 kpc without 
    requiring stacking.  
    We conclusively show that at fixed stellar mass, the stellar profiles of massive 
    galaxies depend on the masses of the their dark matter halos. 
    On average, massive central galaxies in more massive halos have shallower inner 
    stellar mass density profiles (within ${\sim}10$-$20$ kpc) and more prominent 
    outer envelopes. 
    These differences translate into a clear environmental dependence of the 
    mass-size relation: central galaxies in more massive halos have larger 
    $R_{\mathrm{50}}$ at fixed \mtot{}. 
    We further find that the ellipticity of the outer stellar halo also increases 
    with halo mass. 
    Our results are broadly consistent with the two phase formation scenario of 
    massive galaxies in which outer envelopes are built up at late times from the 
    accretion of small systems.

@@

    A key discovery in the last 
    decade has been the dramatic structural transformation of massive quiescent 
    galaxies \citep[e.g.][]{Trujillo2006, vanDokkum2008, Cimatti2008, Damjanov2009, 
    vanderWel2011, Szomoru2012, Patel2013} from $z \approx 2$ to the present day. 
    These observations suggest that the progenitors of $z{\sim} 0$ massive early-type 
    galaxies (ETGs) need to increase their effective radii ($R_{\rm e}$) by a factor 
    of 2--4 over a time span of 10 Gyrs (e.g. \citealt{Newman2012, vdWel2014}). 
    This observational result spurred the development of the ``two-phase'' formation
    scenario for massive ETGs (e.g. \citealt{Oser2010, Oser2012}),
    in which galaxies form a compact central region at $z\sim 2$ through highly 
    dissipative processes (e.g. gas-rich mergers or cold gas-accretion;
    \citealt{Hopkins2008, Dekel2009}). 
    They subsequently assemble extended stellar haloes via dry mergers 
    \citep[e.g,][]{Naab2006, Khochfar2006, Oser2010, Oser2012} which can cause 
    significant size growth at late times. 
    An alternative explanation for size growth, progenitor bias, hypothesizes that 
    larger ETGs were quenched more recently, but this explanation is still under 
    active debate (e.g. \citealt{Newman2012, Carollo2013, Poggianti2013, Belli2015,
    Keating2015, Fagioli2016}).

    There have been multiple observational attempts to test the two-phase 
    formation scenario using galaxies at low redshift, by investigating surface 
    brightness or mass density profiles (e.g. \citealt{Huang2013a, Huang2013b, 
    Oh2017}, optical color gradients (e.g. \citealt{LaBarbera2010, LaBarbera2012}), 
    and stellar population gradients \citep[e.g.,][]{Coccato2010, Coccato2011, 
    Greene2015, Barbosa2016}. Although these observations are generally consistent with the two-phase scenario, this picture still faces a key puzzle: whether or not sizes and profiles of galaxies depend on their mass of their host dark matter halos.

    According to the two-phase formation scenario, the late time growth of massive 
    ETGs is intrinsically tied to the growth of their host dark matter haloes 
    (e.g. \citealt{Leauthaud2012, Behroozi2013, Shankar2013}). 
    Massive ETGs living at the centers of more massive haloes should be subject to 
    higher merger rates compared to ETGs with similar stellar mass but living in less 
    massive dark matter haloes. 
    A natural prediction of this picture is therefore that the structures of 
    massive ETGs should depend on their ``environment''\footnote{There are 
    multiple definitions of ``environment'' in the literature.  
    In this work, we use ``environment'' and halo mass interchangeably.}.
    Theoretical models based on the two-phase formation scenario predict an 
    environmental dependence of the stellar mass--effective radius relation for 
    ETGs (\mstar{}--$R_{\rm e}$; 
    e.g., \citealt{Shen2003, Guo2009}): the sizes of massive ETGs in more massive 
    haloes should be larger than those in less massive haloes at fixed \mstar{} 
    (e.g. \citealt{Shankar2013, Shankar2014}). 
    Yet, solid evidence of this effect at low redshift is still missing 
    (e.g. \citealt{Nair2010, HCompany13}; but also see \citealt{Yoon2017}).

@@

    Testing this prediction requires deep imaging of ETGs in order to probe the 
    accreted component (the stellar halo) which can extend to 100 kpc and 
    beyond \citep[eg.,][]{hscMassiveI}. This is observationally challenging as the surface brightness profiles of 
    massive ETGs decline rapidly with typical values of $\mu > 26.0$ \sb{} in 
    $i$-band at 100 kpc and at $z\sim0.3$. 
    In \citet[][Paper I hereafter]{hscMassiveI}, we showed that deep, 
    multi-band imaging from the Subaru Strategic Program (SSP; \citealt{HSC-SSP,
    HSC-DR1}) using Hyper Suprime-Cam (HSC; \citealt{Miyazaki2012}, 
    Miyazaki in~prep.) allows us to extract robust surface stellar
    mass density (\mden{}) profiles for {\it individual} galaxies with 
    \logms{}$>11.4$ at $0.3 < z < 0.5$ and out to 100 kpc. 
    In Paper I, we characterized the stellar mass profiles of massive ETGs, and showed that there is a large 
    intrinsic scatter in the \mden{} stellar halos of massive galaxies on 100 kpc 
    scales. In this paper, we now investigate whether or not the large scatter in the outer profiles of massive galaxies correlate with halo mass. We conclusively show that the sizes and stellar halos of massive central galaxies depend on dark matter halo mass. 
    In other terms, we reveal the halo mass dependence of the mass-size relation 
    for massive ETGs.

@@

    Magnitudes use the AB system (\citealt{Oke1983}), and are corrected for Galactic 
    extinction using calibrations from \citet{Schlafly11}. We assume $H_0$ = 70~km~s$^{-1}$ Mpc$^{-1}$, ${\Omega}_{\rm m}=0.3$, 
    and ${\Omega}_{\rm \Lambda}=0.7$.
    Stellar mass is denoted \mstar{} and has been derived using a Chabrier Initial Mass 
    Function (IMF; \citealt{Chabrier2003}). Halo mass is defined as 
    $M_{\rm 200b}\equiv M(<r_{\rm 200b})=200\bar{\rho} 
    \frac{4}{3}\pi r_{\rm 200b}^3$ where $r_{\rm 200b}$
    is the radius at which the mean interior density is equal to 200 times
    the mean matter density ($\bar{\rho}$). 
    As in \citet{hscMassiveI}, we do not attempt to decompose or distinguish any  
    potential ``intra-cluster'' 
    component (ICL; e.g. \citealt{Carlberg1997, Lin2004, Gonzalez2005, Mihos2005}).

@@

%% ------------------------------------------------------------------------------------ %% 
\subsection{The Role of Environment in the ``Two--Phase'' Formation Scenario}

\alexie{This section does not have a good logical flow and needs more work and clarity. Ideas are ``choppy''}.

    Our results are broadly consistent with the ``two-phase'' formation theory for massive central galaxies. According to this scenario, intense dissipative processes at $z > 2$ are responsible for 
    the formation of \textit{in situ} stars in massive central galaxies. In the second phase, star-formation is quenched which limits in-situ star formation and growth becomes dominated by non-dissipative mergers of satellite galaxies. Dry minor mergers are efficient at depositing \textit{ex situ} stars in the outskirts of massive galaxies (e.g. \citealt{Oogi2013, Bedorf2013}). Because central galaxies in massive haloes should experience more mergers compared to those in less massive halos, we expect the stellar mass profiles (and sizes) of massive central galaxies to exhibit dependencies on host halo mass.

    \citet{Shankar2014} studied the environment dependence of galaxy size using 
    a series of hierarchical semi-analytic models. They predict that at fixed stellar mass, the median size of central galaxies should increase strongly with halo mass. The major-merger rate increases slightly with halo mass (mass ratio $>1$:3; e.g. \citealt{Hirschmann2013}). The minor-merger rate, however, depends strongly on the dynamical friction time. If the dynamical friction time is short (e.g. \citealt{Newman2012}), then centrals of more massive haloes can have up to four times more minor mergers (1:100--1:3) \alexie{Compare to what? Need to give the halo mass numbers here}. This increase in the minor-merger rate as a function of halo mass leads to a halo-mass dependence in the mass-size relation. The prediction from \citet{Shankar2014} have been confirmed by \citet{Yoon2017} using the semi-analytic model from \citet{Guo2011}.  Our results are broadly consistent with these predictions. Indeed, we find that the outer profiles of massive galaxies ($r>50$ kpc) are the regions that display the strongest variations as a function of halo mass.

    %Whether or not level of difference that we observe is fully consistent with predictions from semi-analytic or hydrodynamic models of galaxy formation is work in progress that will be presented in a follow-up paper.

@@

%% ------------------------------------------------------------------------------------ %% 
\subsection{Comparison with Previous Work}

@@

    Previous work on this topic has sometimes found the mass-size relation to be independent of halo mass (e.g. \citealt{Nair2010, Maltby2010, Cappellari2013, HCompany13}). There are several reasons that might explain why we reach the opposite conclusion compared to these studies: 1) the impact of imaging depth, 2) the method to estimate galaxy ``size'',  the range in  \mstar{} and \mhalo{}.

    \citet{Cerbrian2014} find that ETGs with \logms{}$>11.5$ are slightly larger in 
    more massive haloes, but they also shows that this trend is reversed 
    at lower \mstar{}. \citet{Yoon2017} present a careful study of a large sample of $z\sim0.1$ SDSS ETGs using  
    a non-parametric method and recover a clear environment dependence of mass-size 
    relation at \logms{}$>11.2$. They conclude that, at the same \mstar{}, massive ETGs from dense environment 
    are 20--40\% larger compared to those in underdense environment -- these results are in broad agreement with those presented here.
    \alexie{Need a sentence abbout the new Mike Hudson paper here}

    While there now appears to be an emerging consensus that the high mass end of the mass-size relation does vary with halo mass, the situation at higher redshift is more complicated. Certain studies find an environment dependence of the mass--size relation for massive 
    quiescent or early-type galaxies that is qualitatively consistent with low redshift studies (e.g. \citealt{Papovich2012, Lani2013, Strazzullo2013, 
    Delaye2014}). However, there are also observations that suggest otherwise (e.g. 
    \citealt{Rettura2010, Raichoor2012, Kelkar2015, Allen2015}). These differences might be explained by varying sample sizes,  special environments
    (high-$z$ massive cluster), or the impact of massive star-forming galaxies. However, these yet un-reconciled differences, also underscore the  intrinsic difficulty of measuring stellar masses and sizes 
    for high-$z$ galaxies (e.g. \cite{Price2017}). Instead of using the mass--size relation to compare the high and the low redshift universe, a more robust approach would be to directly compare rest-frame surface brightness 
    or stellar mass density profiles (e.g. \citealt{Szomoru2012, Patel2013, Buitrago2017, Hill2017}).

    \subsection{Towards Consistent Size Definitions}

          Semi-analytic models and hydrodynamic simulations often have difficulties reproducing the  mass--size relation of massive galaxies. Galaxy sizes are sensitive to many different physical processes (star-formation, feedback, mergers) and matching the galaxy stellar-mass function does not automatically guarantee a match to the mass-size relation. Furthermore, while some effort has been made to use consistent size definitions (cite Bahamas paper), more often than not, comparisons of the mass-size relation often do not use consistent size definitions, or only perform crude size conversions (give an example here? 3d radius and 2 radius). For examples, observers often quote ``sizes" which are the effective radius of an elliptical sersic fit. This is a 2d size  corresponding to the major axis of the ellipse that contains 50\% of the light. Simulations on the other hand, often employ sizes that correspond to xxx. In this paper, we quote masses measured within elliptical apertures of fixed physical size. We argue that this approach will allow for a more straightforward comparison between observations and theoretical predictions. In follow-up work, we will present comparisons between our observations and predictions from hydrodynamic simulations using the  using the \mtot{}--\minn{} plane instead of the mass-size relation.
